---
title: "Pond clusters"
author: "Luwen Wan"
date: "2024-06-18"
output:
  html_document:
    self_contained: true
editor_options: 
  chunk_output_type: console
---


```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
rm(list = ls())
library(sf)
library(dplyr)
library(ggplot2)
library(cowplot)
library(mgcv)
library(car)
library(ggpubr)
library(gridExtra)
library(GGally)
library(e1071)
library(tidyr)

# Assuming `gam_pond_simplified` is your fitted GAM model
outdir <- "H:/My Drive/Projects/Beavers/Beaver_EarthEngine/RFanalysis_clusters/outputs"
outfig <- "H:/My Drive/Manuscript/BeavePond_westernUS/Figs"
```

```{r}
load(paste0(outdir,"/merged_pond_var_update.RData"))

names(merged_pond_var_update)
merged_pond_var <- merged_pond_var_update # in seek of convenience 

names(merged_pond_var)

# change name 
names(merged_pond_var) <- c("ClusterID","ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength","PondArea")

merged_pond_var_select <- merged_pond_var[,c("ClusterID","WoodyHeight","SPImedian","DamLength","PondArea")]

```

# 1.  Variable data checking 
## 1.1 sixteen predictors distribution 
```{r}
predictors_all <- merged_pond_var[,c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength")]

# wide to long format 
predictors_all_long <- predictors_all %>%
  pivot_longer(
    cols = c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength"),  
    names_to = c("Variable"),  # New column names
    values_to = "Value"  # Name of the new value column
  )

# Manually set the order of the facets
predictors_all_long$Variable <- factor(predictors_all_long$Variable, 
                                       levels = c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength"))

# remove cluster ID
names(merged_pond_var)
merged_pond_allx_y <- merged_pond_var[,c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength","PondArea")]

```

# 2. VIF (Variance Inflation Factor)
## 2.1 VIF, all < 10
necessary for  both linear and non-linear models when your goal is to explore which factors are impacting ponding size and interpret individual predictors.
```{r}
# VIF is less critical if goal is pure prediction 
names(merged_pond_allx_y)

merged_pond_allx_y$log_PondArea <- log(merged_pond_allx_y$PondArea) # natural logarithms
names(merged_pond_allx_y)

lm_model <- lm(PondArea ~ ., data = merged_pond_allx_y)
vif_values <- vif(lm_model)
vif_values_df <- as.data.frame(vif_values)
print(vif_values)

# unselect pond area as we use log of pond area
merged_pond_allx_logY <- merged_pond_allx_y[,c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength","log_PondArea")]
lm_model_log <- lm(log_PondArea ~ ., data = merged_pond_allx_logY)
vif_values_log <- vif(lm_model_log)
vif_values_log <- as.data.frame(vif_values_log)
```

# 3. normalization 
## 3.1 min-max normalization (0-100) - allow more intuitive comparision about which factors is more important 
```{r}
# Use scaled predictors if their magnitudes differ significantly.
# Define the min-max normalization function to scale data to 0-100
normalize_0_100 <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)) * 100)
}

# Separate the response variable and predictors
names(merged_pond_allx_logY)
response <- merged_pond_allx_logY$log_PondArea
predictors <- merged_pond_allx_logY[, !(names(merged_pond_allx_logY) %in% "log_PondArea")]
names(predictors)

# Apply the min-max normalization to scale the predictors to 0-100
scaled_predictors_0_100 <- as.data.frame(lapply(predictors, normalize_0_100))

# Combine the scaled predictors with the original response variable
merged_pond_allx_logY_scaled_0_100 <- cbind(scaled_predictors_0_100, log_PondArea = response)

# View the resulting dataframe
names(merged_pond_allx_logY_scaled_0_100)
head(merged_pond_allx_logY_scaled_0_100)

# for convenience 
pondFinal <- merged_pond_allx_logY_scaled_0_100

sum(is.na(pondFinal))  # Should be 0 after na.omit if previously handled

```

# 4. Pond Generalized Additive Models (GAM) 
## 4.1 Model 1a - full pond model 
```{r}
# Fit the refined GAM without the interaction term
names(pondFinal)
gam_pond <- gam(
  log_PondArea ~ 
    s(ClayMedian) + 
    s(ThetasMedian) + 
    s(WoodyCover) + 
    s(WoodyHeight) + 
    s(FlowlineLength) + 
    s(MainChannelSlope)+
    s(ValleyBottWidth) + 
    s(UPA) + 
    s(CTImax) + 
    s(SPImedian) + 
    s(STImedian) + 
    s(SummerTempMax) + 
    s(SpringPrecip) + 
    s(AnnualPrecip) + 
    s(SWEMarchMax) + 
    s(DamLength),
  data = pondFinal,
  method = "REML"
)

# Summarize the refined model
summary(gam_pond)
```

## 4.2 model 1b - simplified pond model 
```{r}
# Simplifying a GAM by retaining only the significant factors helps improve interpretability while maintaining predictive power. 
# Refit the GAM with significant predictors
gam_pond_simplified <- gam(
  log_PondArea ~ 
    s(WoodyHeight) + 
    s(DamLength) + 
    s(SPImedian),
  data = pondFinal,
  method = "REML"
)

# Summarize the refined model
summary(gam_pond_simplified)
# plot(gam_pond_simplified, pages = 1, se = TRUE)

AIC(gam_pond, gam_pond_simplified) 


# Plotting all smooth terms and residuals
# par(mfrow = c(2, 2))
# plot(gam_pond_simplified, pages = 1, shade = TRUE, seWithMean = TRUE)

# gam.check(gam_pond_simplified)
# Your model has successfully converged using the REML method with the Outer Newton optimizer. The small gradient values and positive definite Hessian indicate a stable and reliable solution. The full model rank confirms that all predictors contribute uniquely without redundancy.

predicted <- predict(gam_pond_simplified, newdata = pondFinal, type = "response")
# plot(gam_pond_simplified, pages = 1, se = TRUE)
```

### 4.2.0 importance order 
```{r}
# Extract the summary from the model
gam_pond_simplified_summary <- summary(gam_pond_simplified)

# Extract the relevant data for each smooth term
edf <- gam_pond_simplified_summary$s.table[, "edf"]
f_value <- gam_pond_simplified_summary$s.table[, "F"]
p_value <- gam_pond_simplified_summary$s.table[, "p-value"]

# Generate significance levels based on p-values
significance <- ifelse(p_value < 0.001, "***", 
                       ifelse(p_value < 0.01, "**", 
                              ifelse(p_value < 0.05, "*", 
                                     ifelse(p_value < 0.1, ".", " "))))

# Create a data frame with predictor names, EDF, F-values, and significance levels
importance_df_pond  <- data.frame(
  Predictor = rownames(gam_pond_simplified_summary$s.table),
  F_value = f_value,
  EDF = edf,
  Significance = significance
)

pond_order <- ggplot(importance_df_pond, aes(x = reorder(Predictor, F_value), y = F_value)) +
  geom_bar(stat = "identity", fill = "#0070FF", width = 0.5) +
  coord_flip() +
  geom_text(aes(label = paste("EDF =", round(EDF, 2), Significance)), hjust = 1, size = 3) +
  labs(title = "(a) Factor contributions to log-transformed pond area", 
     # x = "Predictor", 
     y = "F-value") +
  theme_bw(base_size = 12) +
  scale_y_continuous(trans = "log1p") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.y = element_blank())

```
 
## 4.3 model 1c - reduced pond model 
```{r}
names(pondFinal)
# Fit the refined GAM without the interaction term
gam_model_noDam <- gam(
  log_PondArea ~ 
    s(ClayMedian) + 
    s(ThetasMedian) + 
    s(WoodyCover) + 
    s(WoodyHeight) + 
    s(FlowlineLength) + 
    s(MainChannelSlope) + 
    s(ValleyBottWidth) + 
    s(UPA) + 
    s(CTImax) + 
    s(SPImedian) + 
    s(STImedian) + 
    s(SummerTempMax) + 
    s(SpringPrecip) + 
    s(AnnualPrecip) + 
    s(SWEMarchMax),
  data = pondFinal,
  method = "REML"
)

# Summarize the refined model
summary(gam_model_noDam)

```

# 5. Dam GAM 
```{r}
# log dam length 
names(merged_pond_allx_y)
# Calculate skewness for DamLenMcluster
skewness_value <- skewness(merged_pond_allx_y$DamLength, na.rm = TRUE)
print(paste("Skewness of Dam Length: ", skewness_value))
merged_pond_allx_y$log_DamLength <- log(merged_pond_allx_y$DamLength)

# Separate the response variable and predictors
response <- merged_pond_allx_y$log_DamLength
names(merged_pond_allx_y)
predictors <- merged_pond_allx_y[, !(names(merged_pond_allx_y) %in% c("DamLength","log_DamLength",
                                                                      "PondArea","log_PondArea"))]
names(predictors)
# Apply the min-max normalization to scale the predictors to 0-100
scaled_predictors_0_100 <- as.data.frame(lapply(predictors, normalize_0_100))

# Combine the scaled predictors with the original response variable
merged_pond_allx_logY_scaled_0_100 <- cbind(scaled_predictors_0_100, log_DamLength = response)

# View the resulting dataframe
head(merged_pond_allx_logY_scaled_0_100)

  # # Linear regression using scaled predictors
  # model_normalize <- lm(log_DamLength ~ ., data = merged_pond_allx_logY_scaled_0_100)
  # 
  # # Summary of the model
  # summary(model_normalize)

# Fit the GAM with log-transformed dam length
names(merged_pond_allx_logY_scaled_0_100)

# for convenience 
damFinal <- merged_pond_allx_logY_scaled_0_100

sum(is.na(damFinal))  # Should be 0 after na.omit if previously handled
```

## 5.1 model2a - full dam gam
```{r}
# linear 
# Linear regression using scaled predictors
lm_dam <- lm(log_DamLength ~ ., data = damFinal)

# Summary of the model
summary(lm_dam)

# gam 
gam_dam <- gam(
  log_DamLength ~ 
    s(ClayMedian) + 
    s(ThetasMedian) + 
    s(WoodyCover) + 
    s(WoodyHeight) + 
    s(FlowlineLength) + 
    s(MainChannelSlope) + 
    s(ValleyBottWidth) + 
    s(UPA) + 
    s(CTImax) + 
    s(SPImedian) + 
    s(STImedian) + 
    s(SummerTempMax) + 
    s(SpringPrecip) + 
    s(AnnualPrecip) + 
    s(SWEMarchMax),
  data = damFinal,
  method = "REML"
)

# Summarize the transformed model
summary(gam_dam)

# AIC and BIC
AIC(lm_dam, gam_dam) # gam should be chosen 
BIC(lm_dam, gam_dam) # difference is 1 BIC has a stronger emphasis on simplicity.
```

## 5.2 model2b - simplified dam gam 
```{r}
# Fit a Generalized Additive Model (GAM) with log-transformed dam length
# using significant predictors
gam_dam_simplified <- gam(
  log_DamLength ~ 
    s(ClayMedian) + 
    s(FlowlineLength) + 
    s(MainChannelSlope) + 
    s(ValleyBottWidth) + 
    s(UPA) + 
    s(CTImax) + 
    s(AnnualPrecip),
  data = damFinal,
  method = "REML"
)

# Summarize the transformed model
summary(gam_dam_simplified)
```


## 5.3 model 2c- interactive dam GAM 
### 5.3.1 test 
```{r}
model_data <- model.frame(gam_dam_simplified)
summary(gam_dam_simplified)

gam_test <- gam(
  log_DamLength ~ 
    s(ClayMedian) + 
    s(FlowlineLength) + 
    ti(ClayMedian,FlowlineLength),
  family = gaussian(), 
  method = "REML", 
  data = damFinal
)

# Print model summary
summary(gam_test)

# Print all terms in s.table
print(rownames(summary(gam_test)$s.table))

# Extract interaction term values manually
interaction_index <- which(rownames(summary(gam_test)$s.table) == "ti(ClayMedian,FlowlineLength)")

if (length(interaction_index) > 0) {
  interaction_row <- summary(gam_test)$s.table[interaction_index, ]
  print(interaction_row)  # Check the actual values for edf, F, and p-value
} else {
  print("Interaction term not found in model summary.")
}

```

### 5.3.2 Loop Through Interactions
### 5.3.3 gam final with significant (p_value < 0.1) interactions 
```{r}
gam_final <- gam(
  log_DamLength ~ 
    s(ClayMedian) + 
    s(FlowlineLength) + 
    s(MainChannelSlope) + 
    s(ValleyBottWidth) + 
    s(UPA) + 
    s(CTImax) + 
    s(AnnualPrecip) + 
    ti(ClayMedian,UPA) + 
    ti(ClayMedian,AnnualPrecip) + 
    ti(FlowlineLength,MainChannelSlope) + 
    ti(ValleyBottWidth,CTImax),
  family = gaussian(), 
  method = "REML", 
  data = damFinal
)

# Print summary
summary(gam_final)

# ti(ClayMedian, UPA) and ti(FlowlineLength, MainChannelSlope) are NOT significant, could be removed 
# Keep the simplified interaction model (gam_final_refined), which includes only significant interactions (ti(ClayMedian, AnnualPrecip) and ti(ValleyBottWidth, CTImax)).

interactive_dam_gam <- gam(
  log_DamLength ~ 
    s(ClayMedian) + 
    s(FlowlineLength) + 
    s(MainChannelSlope) + 
    s(ValleyBottWidth) + 
    s(UPA) + 
    s(CTImax) + 
    s(AnnualPrecip) + 
    ti(ClayMedian,AnnualPrecip) + 
    ti(ValleyBottWidth,CTImax),
  family = gaussian(), 
  method = "REML", 
  data = damFinal
)

summary(interactive_dam_gam)
```


### 5.3.7 model2c - interactive dam gam - dam order 
```{r}
# Summarize the transformed model
summary(interactive_dam_gam)

# Extract the summary from the model
gam_summary <- summary(interactive_dam_gam)

# Extract the relevant data for each smooth term
edf <- gam_summary$s.table[, "edf"]
f_value <- gam_summary$s.table[, "F"]
p_value <- gam_summary$s.table[, "p-value"]

# Generate significance levels based on p-values
significance <- ifelse(p_value < 0.001, "***", 
                       ifelse(p_value < 0.01, "**", 
                              ifelse(p_value < 0.05, "*", 
                                     ifelse(p_value < 0.1, ".", " "))))

# Create a data frame with predictor names, EDF, F-values, and significance levels
importance_df <- data.frame(
  Predictor = rownames(gam_summary$s.table),
  F_value = f_value,
  EDF = edf,
  Significance = significance
)

# Print the importance dataframe for verification
print(importance_df)

# Plot predictor importance from highest to lowest, with significance levels and EDF
dam_order <- ggplot(importance_df, aes(x = reorder(Predictor, F_value), y = F_value)) +
  geom_bar(stat = "identity", fill = "#ff8f00",width = 0.5) +
  coord_flip() +
  geom_text(aes(label = paste("EDF =", round(EDF, 2), Significance)), hjust = 1, size = 3) +
  labs(title = "(b) Factor contributions to log-transformed dam length", 
     # x = "Predictor", 
     y = "F-value") +
  theme_bw(base_size = 12) +
  scale_y_continuous(trans = "log1p") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.y = element_blank())

```

# 4&5. merge pond and dam order 
```{r}
# pond order from summary(gam_pond_simplified)
# dam order from summary(interactive_dam_gam)

x_limit <- c(0, 35)  # Set a consistent range across both plots
# Aligning plots with discrete scales
pond_order_aligned <- pond_order + 
  coord_flip() +  # Flip coordinates for horizontal bars
  theme(
    plot.margin = unit(c(5, 5, 5, 5), "pt")  # Add consistent plot margins
  )

dam_order_aligned <- dam_order + 
  coord_flip() +  # Flip coordinates for horizontal bars
  theme(
    plot.margin = unit(c(5, 5, 5, 5), "pt")  # Add consistent plot margins
  )


combined_plot <- plot_grid(
  pond_order_aligned, 
  dam_order_aligned, 
  ncol = 1,                 # Arrange in a single column
  rel_heights = c(3, 7),    # Proportional heights
  align = "v"               # Align vertically
)

# Print or save the combined plot
print(combined_plot)

```

```{r}
# Save the plot
# fname <- paste0(outfig, '/var_importance_update.png')
# ggsave(filename = fname, plot = combined_plot, dpi = 600, width = 7, units = 'in')
```
