---
title: "Pond partial effects with crossing"
author: "Luwen Wan"
date: "2024-06-18"
output:
  html_document:
    self_contained: true
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
rm(list = ls())
library(sf)
library(dplyr)
library(ggplot2)
library(cowplot)
library(mgcv)
library(car)
library(ggpubr)
library(gridExtra)
library(GGally)
library(e1071)
library(tidyr)

# Assuming `gam_pond_simplified` is your fitted GAM model
outdir <- "H:/My Drive/Projects/Beavers/Beaver_EarthEngine/RFanalysis_clusters/outputs"
outfig <- "H:/My Drive/Manuscript/BeavePond_westernUS/Figs"
```

```{r}
load(paste0(outdir,"/merged_pond_var_update.RData"))

names(merged_pond_var_update)
merged_pond_var <- merged_pond_var_update # in seek of convenience 

names(merged_pond_var)

# change name 
names(merged_pond_var) <- c("ClusterID","ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength","PondArea")

merged_pond_var_select <- merged_pond_var[,c("ClusterID","WoodyHeight","SPImedian","DamLength","PondArea")]

```

# 1.  Variable data checking 
## 1.1 sixteen predictors distribution 
```{r}
predictors_all <- merged_pond_var[,c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength")]

# wide to long format 
predictors_all_long <- predictors_all %>%
  pivot_longer(
    cols = c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength"),  
    names_to = c("Variable"),  # New column names
    values_to = "Value"  # Name of the new value column
  )

# Manually set the order of the facets
predictors_all_long$Variable <- factor(predictors_all_long$Variable, 
                                       levels = c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength"))

# remove cluster ID
names(merged_pond_var)
merged_pond_allx_y <- merged_pond_var[,c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength","PondArea")]

```

# 2. VIF (Variance Inflation Factor)
## 2.1 VIF, all < 10
necessary for  both linear and non-linear models when your goal is to explore which factors are impacting ponding size and interpret individual predictors.
```{r}
# VIF is less critical if goal is pure prediction 
names(merged_pond_allx_y)

merged_pond_allx_y$log_PondArea <- log(merged_pond_allx_y$PondArea) # natural logarithms
names(merged_pond_allx_y)

lm_model <- lm(PondArea ~ ., data = merged_pond_allx_y)
vif_values <- vif(lm_model)
vif_values_df <- as.data.frame(vif_values)

# unselect pond area as we use log of pond area
merged_pond_allx_logY <- merged_pond_allx_y[,c("ClayMedian","ThetasMedian","WoodyCover","WoodyHeight","FlowlineLength",
                                         "MainChannelSlope","ValleyBottWidth","UPA","CTImax","SPImedian",
                                         "STImedian","SummerTempMax","SpringPrecip","AnnualPrecip","SWEMarchMax","DamLength","log_PondArea")]
lm_model_log <- lm(log_PondArea ~ ., data = merged_pond_allx_logY)
vif_values_log <- vif(lm_model_log)
vif_values_log <- as.data.frame(vif_values_log)
```

# 3. normalization 
## 3.1 min-max normalization (0-100) - allow more intuitive comparision about which factors is more important 
```{r}
# Use scaled predictors if their magnitudes differ significantly.
# Define the min-max normalization function to scale data to 0-100
normalize_0_100 <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)) * 100)
}

# Separate the response variable and predictors
names(merged_pond_allx_logY)
response <- merged_pond_allx_logY$log_PondArea
predictors <- merged_pond_allx_logY[, !(names(merged_pond_allx_logY) %in% "log_PondArea")]
names(predictors)

# Apply the min-max normalization to scale the predictors to 0-100
scaled_predictors_0_100 <- as.data.frame(lapply(predictors, normalize_0_100))

# Combine the scaled predictors with the original response variable
merged_pond_allx_logY_scaled_0_100 <- cbind(scaled_predictors_0_100, log_PondArea = response)

# View the resulting dataframe
names(merged_pond_allx_logY_scaled_0_100)
head(merged_pond_allx_logY_scaled_0_100)

# for convenience 
pondFinal <- merged_pond_allx_logY_scaled_0_100

sum(is.na(pondFinal))  # Should be 0 after na.omit if previously handled

```

# 4. Pond Generalized Additive Models (GAM) 

## 4.2 model 1b - simplified pond model 
```{r}
# Simplifying a GAM by retaining only the significant factors helps improve interpretability while maintaining predictive power. 
# Refit the GAM with significant predictors
gam_pond_simplified <- gam(
  log_PondArea ~ 
    s(WoodyHeight) + 
    s(DamLength) + 
    s(SPImedian),
  data = pondFinal,
  method = "REML"
)

# Summarize the refined model
summary(gam_pond_simplified)
# plot(gam_pond_simplified, pages = 1, se = TRUE)

# Plotting all smooth terms and residuals
# par(mfrow = c(2, 2))
# plot(gam_pond_simplified, pages = 1, shade = TRUE, seWithMean = TRUE)

# gam.check(gam_pond_simplified)
# Your model has successfully converged using the REML method with the Outer Newton optimizer. The small gradient values and positive definite Hessian indicate a stable and reliable solution. The full model rank confirms that all predictors contribute uniquely without redundancy.

predicted <- predict(gam_pond_simplified, newdata = pondFinal, type = "response")
# plot(gam_pond_simplified, pages = 1, se = TRUE)
```



# 6. Partial Effect 
  <!-- ### 6.1.1 PE Plots with scaled value - 3 panels -->
  <!-- ```{r} -->
  <!-- # The y-axis (.estimate) represents the estimated smooth effect of that predictor on the response variable. This effect is the result of applying the smooth term fitted by the Generalized Additive Model (GAM). -->
  
  <!-- # Extract smooth effects for each term -->
  <!-- woody_height_df <- gratia::smooth_estimates(gam_pond_simplified, select = "s(WoodyHeight)") -->
  <!-- dam_length_df <- gratia::smooth_estimates(gam_pond_simplified, select = "s(DamLength)") -->
  <!-- spi_median_df <- gratia::smooth_estimates(gam_pond_simplified, select = "s(SPImedian)") -->
  
  <!-- # Create partial effect plots using ggplot2 and theme_bw -->
  <!-- p1 <- ggplot(woody_height_df, aes(x = WoodyHeight, y = .estimate)) + -->
  <!--   geom_line(color = "blue") + -->
  <!--   geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se), alpha = 0.2, fill = "blue") + -->
  <!--   labs(title = "Partial Effect of Woody Height", x = "Woody Height", y = "Partial Effect") + -->
  <!--   theme_bw() -->
  <!-- p1 -->
  
  <!-- p2 <- ggplot(dam_length_df, aes(x = DamLength, y = .estimate)) + -->
  <!--   geom_line(color = "green") + -->
  <!--   geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se), alpha = 0.2, fill = "green") + -->
  <!--   labs(title = "Partial Effect of Dam Length", x = "Dam Length (m)", y = "Partial Effect") + -->
  <!--   theme_bw() -->
  <!-- p2 -->
  <!-- p3 <- ggplot(spi_median_df, aes(x = SPImedian, y = .estimate)) + -->
  <!--   geom_line(color = "red") + -->
  <!--   geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se), alpha = 0.2, fill = "red") + -->
  <!--   labs(title = "Partial Effect of SPI Median", x = "SPI Median", y = "Partial Effect") + -->
  <!--   theme_bw() -->
  
  <!-- # Display the plots together in a single column layout -->
  <!-- grid.arrange(p1, p2, p3, nrow = 3) -->
  <!-- ``` -->

## 6.1 PE Plots with original value
```{r}
# Extract partial effect data
woody_height_df <- gratia::smooth_estimates(gam_pond_simplified, smooth = "s(WoodyHeight)")
dam_length_df <- gratia::smooth_estimates(gam_pond_simplified, smooth = "s(DamLength)")
spi_median_df <- gratia::smooth_estimates(gam_pond_simplified, smooth = "s(SPImedian)")

# Partial Effect of Dam Length
# Define a function to plot partial effects
plot_partial_effect <- function(df, variable_name, color, title, x_label) {
  ggplot(df, aes(x = .data[[variable_name]], y = .estimate)) +
    geom_line(color = color, linewidth = 1) +
    geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se), alpha = 0.2, fill = color) +
    labs(title = title, x = x_label, y = "Partial Effect") +
    theme_bw()
}

# Generate partial effect plots
p1 <- plot_partial_effect(woody_height_df, "WoodyHeight", "green", "(a) Woody Height", "Woody Height (m)")
p2 <- plot_partial_effect(dam_length_df, "DamLength", "blue", "(b) Dam Length", "Dam Length (m)")
p3 <- plot_partial_effect(spi_median_df, "SPImedian", "red", "(c) SPI Median", "SPI Median")

# Display plots
# grid.arrange(p1, p2, p3, nrow = 1)

# test plot f
predictors <- merged_pond_allx_logY[, c("WoodyHeight", "DamLength", "SPImedian")]

# Correct computation of min and max values
min_values <- sapply(predictors, function(x) min(x, na.rm = TRUE))
max_values <- sapply(predictors, function(x) max(x, na.rm = TRUE))

print(min_values)
print(max_values)

back_transform_0_100 <- function(scaled, min_val, max_val) {
  return ((scaled / 100) * (max_val - min_val) + min_val)
}

# Define colors for each predictor
colors <- c("WoodyHeight" = "green", "DamLength" = "blue", "SPImedian" = "red")
# colors <- c("WoodyHeight" = "#4daf4a", "DamLength" = "#377eb8", "SPImedian" = "#e41a1c")
line_types <- c("dashed", "dashed", "dashed")

# Partial effect plot for DamLength
dam_length_df$DamLength_original <- back_transform_0_100(
  scaled = dam_length_df$DamLength,
  min_val = min_values["DamLength"],
  max_val = max_values["DamLength"]
)
names(dam_length_df)

p1 <- ggplot(dam_length_df, aes(x = DamLength_original, y = .estimate)) +
  geom_line(color = colors["DamLength"],linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = colors["DamLength"]) +
  labs(title = "Partial effects of significant predictors on log-transformed pond area",
       x = "Dam Length (m)") +
   theme_bw(base_size = 10) +
   theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank(),
    axis.title.x = element_text(),
    axis.text = element_text()
  ) +
  annotate("text", x = -Inf, y = Inf, label = "(a)", hjust = -1, vjust = 2, size = 4)

# Partial effect plot for SPImedian
spi_median_df$SPImedian_original <- back_transform_0_100(
  scaled = spi_median_df$SPImedian,
  min_val = min_values["SPImedian"],
  max_val = max_values["SPImedian"]
)

p2 <- ggplot(spi_median_df, aes(x = SPImedian_original, y = .estimate)) +
  geom_line(color = colors["SPImedian"],linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = colors["SPImedian"]) +
  labs(x = "SPI Median") +
  theme_bw(base_size = 10) +
   theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank(),
    axis.title.x = element_text(),
    axis.text = element_text()
  ) +
  annotate("text", x = -Inf, y = Inf, label = "(b)", hjust = -1, vjust = 2, size = 4)

# Partial effect plot for WoodyHeight
woody_height_df$WoodyHeight_original <- back_transform_0_100(
  scaled = woody_height_df$WoodyHeight,
  min_val = min_values["WoodyHeight"],
  max_val = max_values["WoodyHeight"]
)

p3 <- ggplot(woody_height_df, aes(x = WoodyHeight_original, y = .estimate)) +
  geom_line(color = colors["WoodyHeight"],linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = colors["WoodyHeight"]) +
  labs(x = "Woody Height (m)") +
  theme_bw(base_size = 10) +
   theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank(),
    axis.title.x = element_text(),
    axis.text = element_text()
  ) +
  annotate("text", x = -Inf, y = Inf, label = "(c)", hjust = -1, vjust = 2, size = 4)

# Arrange all three plots in a single figure
# combined_pond_plot <- grid.arrange(p1, p2, p3, nrow = 3)
```

## 6.2 Add zero-crossings 
```{r}
# Add a column for the sign of .estimate
woody_height_df$sign <- sign(woody_height_df$.estimate)

# Find rows where the sign changes (zero-crossings)
zero_crossings <- which(diff(woody_height_df$sign) != 0)

# Function to interpolate zero-crossings
find_zero_crossings <- function(x1, x2, y1, y2) {
  return(x1 - y1 * (x2 - x1) / (y2 - y1))
}

# Interpolate zero-crossing values
zero_points_scaled <- mapply(
  find_zero_crossings,
  x1 = woody_height_df$WoodyHeight[zero_crossings],
  x2 = woody_height_df$WoodyHeight[zero_crossings + 1],
  y1 = woody_height_df$.estimate[zero_crossings],
  y2 = woody_height_df$.estimate[zero_crossings + 1]
)

# Back-transform scaled zero-crossing points
zero_points_original <- back_transform_0_100(
  scaled = zero_points_scaled,
  min_val = min_values["WoodyHeight"],
  max_val = max_values["WoodyHeight"]
)

print(zero_points_original)

# Function to find zero-crossings and back-transform
find_zero_crossings_for_predictor <- function(smooth_df, predictor_name) {
  # Add sign column for detecting sign changes
  smooth_df$sign <- sign(smooth_df$.estimate)
  
  # Identify zero-crossings (rows where the sign changes)
  zero_crossings <- which(diff(smooth_df$sign) != 0)
  
  # Interpolate to find exact zero-crossing points (scaled)
  zero_points_scaled <- mapply(
    find_zero_crossings,
    x1 = smooth_df[[predictor_name]][zero_crossings],
    x2 = smooth_df[[predictor_name]][zero_crossings + 1],
    y1 = smooth_df$.estimate[zero_crossings],
    y2 = smooth_df$.estimate[zero_crossings + 1]
  )
  
  # Back-transform to original scale
  zero_points_original <- back_transform_0_100(
    scaled = zero_points_scaled,
    min_val = min_values[predictor_name],
    max_val = max_values[predictor_name]
  )
  
  return(zero_points_original)
}

# List of predictors
predictors <- c("WoodyHeight", "DamLength", "SPImedian")

# Initialize a list to store results
zero_crossings_list <- list()

# Loop through predictors and calculate zero-crossings
for (predictor in predictors) {
  # Extract smooth estimates for the predictor
  smooth_df <- gratia::smooth_estimates(gam_pond_simplified, smooth = paste0("s(", predictor, ")"))
  
  # Add original values to smooth_df
  smooth_df[[paste0(predictor, "_original")]] <- back_transform_0_100(
    scaled = smooth_df[[predictor]],
    min_val = min_values[predictor],
    max_val = max_values[predictor]
  )
  
  # Find zero-crossings and store results
  zero_crossings_list[[predictor]] <- find_zero_crossings_for_predictor(
    smooth_df = smooth_df,
    predictor_name = predictor
  )
}

# Print zero-crossings for all predictors
print(zero_crossings_list)

# Define colors for predictors
colors <- c("WoodyHeight" = "green", "DamLength" = "blue", "SPImedian" = "red")

# Function to limit zero-crossings to a maximum of two
limit_zero_crossings <- function(zero_points, max_crossings = 2) {
  if (length(zero_points) > max_crossings) {
    return(zero_points[1:max_crossings])  # Keep only the first `max_crossings` points
  }
  return(zero_points)  # Return all if fewer than `max_crossings`
}

# Zero-crossings for DamLength
dam_length_zero_points <- limit_zero_crossings(zero_crossings_list$DamLength, max_crossings = 2)

# Plot for DamLength
p1 <- ggplot(dam_length_df, aes(x = DamLength_original, y = .estimate)) +
  geom_line(color = colors["DamLength"], linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se), 
              alpha = 0.2, fill = colors["DamLength"]) +
  # geom_vline(xintercept = dam_length_zero_points, color = "black", linetype = "dashed") +
  annotate("text", x = dam_length_zero_points, y = 0, label = round(dam_length_zero_points, 2), 
           color = "black", vjust = -1, size = 3) +
  labs(title = "Partial effects of significant predictors on log-transformed pond area", 
       x = "Dam Length (m)") +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank(),
    axis.title.x = element_text(),
    axis.text = element_text()
  ) +
  annotate("text", x = -Inf, y = Inf, label = "(a)", hjust = -1, vjust = 2, size = 4)+
  geom_hline(yintercept = 0, color = "black", linetype = "solid")

# Zero-crossings for SPImedian
spi_median_zero_points <- limit_zero_crossings(zero_crossings_list$SPImedian, max_crossings = 2)

# Plot for SPImedian
p2 <- ggplot(spi_median_df, aes(x = SPImedian_original, y = .estimate)) +
  geom_line(color = colors["SPImedian"], linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se), 
              alpha = 0.2, fill = colors["SPImedian"]) +
  # geom_vline(xintercept = spi_median_zero_points, color = "black", linetype = "dashed") +
  annotate("text", x = spi_median_zero_points, y = 0, label = round(spi_median_zero_points, 2), 
           color = "black", vjust = -1, size = 3) +
  labs(x = "SPI Median") +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank(),
    axis.title.x = element_text(),
    axis.text = element_text()
  ) +
  annotate("text", x = -Inf, y = Inf, label = "(b)", hjust = -1, vjust = 2, size = 4)+
  geom_hline(yintercept = 0, color = "black",linetype = "solid")  # Adjust size for boldness

# Zero-crossings for WoodyHeight
woody_height_zero_points <- limit_zero_crossings(zero_crossings_list$WoodyHeight, max_crossings = 2)

# Plot for WoodyHeight
p3 <- ggplot(woody_height_df, aes(x = WoodyHeight_original, y = .estimate)) +
  geom_line(color = colors["WoodyHeight"], linetype = "dashed", linewidth = 1) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se), 
              alpha = 0.2, fill = colors["WoodyHeight"]) +
  # geom_vline(xintercept = woody_height_zero_points, color = "black", linetype = "dashed") +
  annotate("text", x = woody_height_zero_points, y = 0, label = round(woody_height_zero_points, 2), 
           color = "black", vjust = -1, size = 3) +
  labs(x = "Woody Height (m)") +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.y = element_blank(),
    axis.title.x = element_text(),
    axis.text = element_text()
  ) +
  annotate("text", x = -Inf, y = Inf, label = "(c)", hjust = -1, vjust = 2, size = 4)+
  geom_hline(yintercept = 0, color = "black", linetype = "solid")

# Combine all plots into one figure
combined_pond_plot <- plot_grid(
  p1, p2, p3, 
  # labels = c("(a)", "(b)", "(c)"),  # Add labels for subplots
  ncol = 1,  # Arrange plots in a single column
  align = "v",  # Align vertically
  label_size = 14  # Adjust label size
)

combined_pond_plot
```

```{r}
# Save the plot
# fname <- paste0(outfig, '/pond_partial_effects_crossing.png')
# ggsave(filename = fname, plot = combined_pond_plot, dpi = 600, width = 6.5, units = 'in')
```

